Individual and Differential Harm in Redistricting

### 2022-08-27 paper/harm.Rmd
editing outside intro

Our definition of harm is limited in two notable regards.
First, it considers only individuals who vote: people who abstain from voting have a harm of zero.
Empirically, this changes nothing, but limits the measure in a theoretical sense.
People who do not voter express no formal preference, even if they have informal preferences.
The natural theoretical extension of this definition is to then separate individual preferences from vote choice. 
Naturally, extending this approach requires auxiliary data to implement, such as an absurdly large-scale survey, as candidate preferences for non-voters are not directly inferred from election returns.
As surveys are typically benchmarked against electoral outcomes, we do not endorse this type of extension.

@@

As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
In general, we are interested in outcomes over multiple elections and a much larger set of counterfactual plans.
Ideally, we would average individual harm (1) over a set of future elections under the enacted districting plan and (2) over a set of hypothetical districting plans which are consistent with districting criteria.
We could even consider a set of individuals with some shared characteristic(s).

@@

Election results present a different set of modeling challenges, given the large number of factors that influence the outcome, and the general unpredictability of future events.
We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting Estimating differential racial harm in the two-party setting can be done by first generating ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
Multiparty or primary elections pose additional modeling challenges.
In Section \ref{sec:boston} we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.

@@

The modern analysis of racial gerrymandering and vote dilution are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th and 15 Amendments.
Section 2 of the VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
Under the current framework established by @1986gingles, to do so, plaintiffs must prove that (a) the minority group is large and geographically compact enough to form a majority in some hypothetical district, (b) the minority group is "politically cohesive," and (c) the majority group can and does vote to deny the minority group its choice of candidate [@hood2018legal] (the so-called *Gingles* criteria).
Criteria (a) is generally proven through a single example districting plan, while criteria (b) and (c) are shown using the statistical estimates of vote preferences and turnout by race.

@@

Taken together, these findings suggest that despite the dominance of the Democratic party in city politics, and the shared preferences between certain groups of Black and White voters, race-aware districting still serves an important role in empowering Black voters across the city to elect their chosen candidates, and that a switch to race-neutral boundaries would have a material impact on Black voting power.
Further study could focus on the racial impact of the four at-large members in comparison to a system with 13 districts and no at-large members.

>>>>

Our definition of harm is limited in two notable regards.
First, it considers only individuals who vote: people who abstain from voting have a harm of zero.
People who do not voter express no formal preference in electoral outcomes, even if they have informal preferences.
The natural theoretical extension of this definition is to then separate individual preferences from vote choice. 
The latter defines election winners, but the former provides a more-broad harm measure.
Naturally, extending this approach requires auxiliary data to implement, such as an absurdly large-scale survey, as candidate preferences for non-voters are not directly inferred from election returns.
As surveys are typically benchmarked against electoral outcomes, we do not endorse this type of extension.

@@

As defined above, $h_i(v,q,\tilde q)$ captures harm in a single election and relative to one counterfactual districting plan $\tilde q$.
In general, we are interested in outcomes over multiple elections and a much larger set of counterfactual plans.
Ideally, we would average individual harm (1) over a set of future elections under the enacted districting plan and (2) over a set of hypothetical districting plans which are consistent with districting criteria.
We could even consider a set of individuals with some shared characteristics.

@@

Election results present a different set of modeling challenges, given the large number of factors that influence the outcome, and the general unpredictability of future events.
We limit our discussion here to the case of estimating harm and differential partisan harm in a two-party setting. 
Estimating differential racial harm in the two-party setting can be done by first generating ecological inferences or survey estimates of party preference by race, and then applying methods analogous to those outlined here.
Multiparty or primary elections pose additional modeling challenges.
In Section \ref{sec:boston} we consider a spatial-racial model for primary elections with multiple candidates, to illustrate the application of our framework to these more complicated cases.

@@

The modern analysis of vote dilution and racial gerrymandering are inseparable from the Voting Rights Act of 1965 (VRA) and the 14th and 15 Amendments.
Section 2 of the VRA enables citizens to challenge districting plans in court for diluting the voting power of racial minorities.
Under the current framework established by @1986gingles, to do so, plaintiffs must prove that (a) the minority group is large and geographically compact enough to form a majority in some hypothetical district, (b) the minority group is "politically cohesive," and (c) the majority group can and does vote to deny the minority group its choice of candidate [@hood2018legal] (the so-called *Gingles* criteria).
Criteria (a) is generally proven through a single example districting plan, while criteria (b) and (c) are shown using the statistical estimates of vote preferences and turnout by race.

@@

Taken together, these findings suggest that despite the dominance of the Democratic party in city politics, and the shared preferences between certain groups of Black and White voters, race-aware districting still serves an important role in empowering Black voters across the city to elect their chosen candidates.
A switch to race-neutral boundaries would have a material impact on Black voting power.
Further study could focus on the racial impact of the four at-large members in comparison to a system with 13 districts and no at-large members.
